---
title:    "Final Project"
subtitle: "STAT 244"
author:   "Prashikshya Pokhrel"
date:     "2025-05-12"
format:   pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE) 
library(readr) 
library(purrr) 
library(dplyr) 
library(ggplot2) 
library(gridExtra) 
library(mosaic) 
library(gmodels) 
library(Sleuth3) 
library(tidyverse) 
library(tidymodels) 
library(readxl)
```

### 1. Describe your data set in 1 - 2 sentences.

Answer: The GrowthDJ dataset is a cross-country growth‐regression sample compiled by Durlauf & Johnson (1995). It contains 121 observations on 10 variables, covering the period 1960–1985.

### 2. What is the source of your data set? Include a link to where you got it here.

**Note.** If you would like to collect survey data from classmates or peers via a Google form, please let me know right away so we can set this up for you. In the meantime, create the Google Form and decide what information you want to collect.

The following questions allow us to determine if your proposed data set meets the project criterion:

Ansewr: <https://vincentarelbundock.github.io/Rdatasets/doc/AER/GrowthDJ.html>

### 3. Load your data set using the code chunk below.

```{r}
df = read.csv("GrowthDJ.csv") 
head(df) 
nrow(df) 
ncol(df) 
# df after cleaning 
df_clean <- na.omit(df) 
head(df_clean) 
nrow(df_clean) 
ncol(df_clean)
```

### 4. How many rows does your data set have? There should be at least 30 rows.

Answer: My dataset had 121 rows initially but after cleaning and dropping the empty rows, now it has 100 rows and 11 columns including rownames.

### 5. How many quantitative variables are in your data set? There should be at least 3. List them here and briefly describe what they represent **and** their units.

Answer: If there are more than (say) 5 quantitative variables in your data set, you only need to list the ones you would like to consider. You can of course decide to consider other variables later; your choices in the checkpoints are not binding.

Answer: There are 8 quantitative variables and I considered choosing 4 of them as of right now. Key details:

gdpgrowth (numerical)

-   Average growth rate of per capita GDP from 1960 to 1985 (in percent).

popgrowth (numerical)

-   Average growth rate of working-age population 1960 to 1985 (in percent).

invest (numerical)

-   Average ratio of investment (including Government Investment) to GDP from 1960 to 1985 (in percent).

school (numerical)

-   Average fraction of working-age population enrolled in secondary school from 1960 to 1985 (in percent).

### 6. How many categorical variables does your data set have? There should be at least 1. List the categorical variable(s) here along with their corresponding categories.

Answer: If there are more than (say) 3 categorical variables in your data set, you only need to list the ones you would like to consider. You can of course decide to consider other variables later; your choices in the checkpoints are not binding.

There are 3 categorical variables and I considered choosing 2 of them as of right now. Key details:

oil (categorical)

-   Factor. Is the country an oil-producing country?

oecd (categorical)

-   Factor. Is the country a member of the OECD?

### 7. What variables in your data set are you interested in? Are they quantitative or categorical? If quantitative, provide the units. If categorical, list or describe the categories.

Answer: I am interested in seeing if the oil producing status of a country will affect the per capita gdp growth of that country. For this I am interested in oil variable which has two category "yes" and "no" (categorical variable), gdpgrowth (numerical variable), school (numerical variable) and invest (numerical variable) to also see if school or investment can also confound the gdp growth of a country even if the country produces an oil?

### 8. What does one observational unit (row) represent in your data set?

Answer: Each row in my data set corresponds to a single “country‐year” observation—that is, one country in one specific year.

### 9. How was the sample obtained? If you can find what the sampling method was, include that information. Also include what year the data was collected

##### a. If any of this information isn’t available, state that.

Answer: The GrowthDJ dataset covers 121 countries for which Penn World Table 4.0 and World Bank (World Development Report) data were available, averaged over 1960–1985. Sampling method has not been explicitly stated in the data set.

##### b. How did you access the data? (Include a hyperlink, and the date + time you most recently accessed the data.)

Answer: <https://vincentarelbundock.github.io/Rdatasets/doc/AER/GrowthDJ.html>, 4/30/2025 10:08 PM

### 10. Do your quantitative variables have the correct types (e.g., double for decimals, integer for whole numbers, etc.). If not, set them to be the correct type. You can check the type of a variable using the class() function.

```{r}
class(df_clean$gdpgrowth) 
class(df_clean$invest) 
class(df_clean$school) 
```

Answer: All my quantitative variables have the correct types.

### 11. Do your categorical variables have the type factor? If not, give them this type to tell R that they are categorical.

```{r}
class(df_clean$oil) 
df_clean$oil <- as.factor(df_clean$oil)
```

### 12. Are there any missing values for your variables listed in (9)? If so, filter them out. Note: Some data sets use empty strings “” to denote missing values, so check for these as well.

```{r}
df_clean <- df_clean %>% 
  filter(!is.na(oil) & !is.na(gdpgrowth) & !is.na(invest) & !is.na(school))   
nrow(df_clean)
```

### 13. How many observational units (rows) do you have after the previous step?

Answer: There are 100 observational units (rows) after cleaning the data set.

### 14. Does the source of data say that certain data points are missing? (For example, in NHANES we are told that the variable Height is measured only for participants aged 2 years or older, meaning Height will have NA values in rows corresponding to participants under 2.) If not available in the data description, why do you think some data points are missing? Is the mechanism for missingness completely random, or could there be something systematic which leads to greater rates of missingness? It’s okay if you’re not sure how to answer this question — just do your best to think about it.

Answer: The GrowthDJ data don’t specify this rule, so any NAs (e.g. in school, literacy60 or invest) likely reflect gaps in historical reporting especially for low-income or non-OECD countries with weaker statistical systems.

### 15. Provide any numerical summaries that are relevant to your research question. These are highly dependent on your question, but some ideas are:

##### For one quantitative variable, some relevant numerical summaries involve:

##### a. One measure of center (e.g., mean, median);

```{r}
summary(df_clean$gdpgrowth) 
```

##### b. One measure of spread (e.g., standard deviation, variance, IQR, a minimum/maximum);

```{r}
sd(df_clean$gdpgrowth) 
```

##### c. An overall count (how many observations do you have?)

```{r}
nrow(df_clean) 
```

##### For one categorical variable, a relevant numerical summary might be:

##### a. The \# of observations in each category (i.e., use count());

```{r}
table(df_clean$oil)
```

##### b. The proportion (or percentage) of observations in each category.

```{r}
prop <- prop.table(table(df_clean$oil)) 
prop
```

### 16. Provide any visualizations that are relevant to your research question. Chapter 2 in this online text is a good resource for creating different plots.

##### The choice of visualization can be highly dependent on your question, but some ideas are:

##### a. A visual summary of your (quantitative) outcome of interest:

##### i. Examples: a density plot, a box plot, a histogram, a relative frequency histogram, etc.

```{r}
ggplot(df_clean, aes(x = gdpgrowth)) + geom_histogram(binwidth = 1, fill = "white", color = "black") + labs(title = "Histogram of Per-Capita GDP Growth", x = "GDP Growth", y = "Count") 
```

##### b. A visual summary of your predictor(s) of interest:

##### i. For one categorical predictor, you could do a bar chart, mosaic plot, etc.

```{r}
ggplot(df_clean, aes(x = oil)) +geom_bar(color="black", fill="pink") +labs(title = "Count of Oil-Producing Status",x= "Oil-Producing Status", y= "Count") + theme_minimal() 
```

##### c. A visual summary of the relationship between one quantitative variable and one or more different variables:

##### i. Ideas: scatter plots (maybe with points colored by a category of some categorical variable), side-by-side histograms, side-by-side box plots, side-by-side histograms, overlaid density plots, etc.

```{r}
ggplot(df_clean, aes(x = invest, y = gdpgrowth, color = oil)) + geom_point(alpha = 0.7) + labs(title = "GDP Growth vs. Investment Ratio, by Oil Status",x= " Average ratio of Investment in percentage",y = "Per-Capita GDP Growth",color = "Oil Producer") 
```

```{r}
ggplot(df_clean, aes(x = gdpgrowth, color = oil)) +geom_density(size = 1) +labs(title = "Density of GDP Growth by Oil Status",x = "Per-Capita GDP Growth",y = "Density")  
```

# Cross Validation (CV)

### 1. Write a brief introduction to cross validation which includes relevant mathematical notation.

Answer: Cross validation is a statistical method used in machine learning to evaluate how well a model performs in an unseen or independent data set. It involves dividing the data set into multiple folds or subsets, using one of these folds as a validation set and training the model on the remaining folds. This process is then repeated multiple times each time using different fold as the validation set. The main purpose of cross validation is to prevent over fitting which occurs when a model is trained too well on the training data and performs poorly on new, unseen data

$\mathrm{CV}(k) = \frac{1}{k}\sum_{j=1}^k \mathrm{MAE}_j$

### 2. What is the goal of cross-validation? (Hint: Think about over-fitting, along with social/ethical considerations)

Answer: The goal of cross validation is to ensure that a model generalizes well beyond the data on which it was trained by repeatedly holding out different subsets fir testing, it guards against over fitting and produces an approximately unbiased estimate of out of sample error. This process not only helps in selecting and tuning models that truly capture underlying patterns rather than noise, but also promotes robustness and fairness: by validating across different folds.

### 3. What linear models are you considering based on your research question? Pick at least two models to compare.

Answer: I am considering two models. One model which is unadjusted model (model_1) which doesn't contain confounders and another model which is adjusted model (model_2) that includes the confounders.

```{r}
# Step 1: Model specification  
lm_spec <- linear_reg() %>%  
  # Output Y is quantitative  
  set_mode("regression") %>% 
  # Want regression to be linear  
  set_engine("lm")
```

```{r}
# Unadjusted model  
model_1 <- lm_spec %>%       
  fit(gdpgrowth ~ oil, data = df_clean)
```

```{r}
# Adjusted model  
model_2 <- lm_spec %>%       
  fit(gdpgrowth ~ oil + school + invest, data = df_clean)
```

### 4. Explain how you divided your data into its test set and training set.

Answer: I split the data into two parts –a training set (80% of the observations) and a test set (the remaining 20%) so that we could both tune our models and then evaluate their performance on truly unseen data.

```{r}
split <- initial_split(df_clean, prop = 0.8, strata = "oil")   
# After splitting: 
train_data <- training(split)  
test_data  <- testing(split)
```

### 5. State which error metric you are using (MAE or MSE) and give its formal mathematical definition. Why did you choose this error metric? What are the advantages/disadvantages of using it?

Answer: I will be using MAE. The formal mathematical definition is:

$\mathrm{MAE} = \displaystyle\frac{1}{n}\sum_{i=1}^n |y_i - \hat y_i|$

I chose this error metric because it produces an error metric in the same units as the outcome which makes it immediately interpretable and it is also less sensitive to large outliers than the other error metric like MSE. The main advantage of MAE is the robustness to outliers. Since errors aren't squared a few extreme misses don't dominate the metric. Another advantage is that it's easy to compute and understand which doesn't require additional transformations.

However, it does have some drawbacks like MAE treats every error the same so one huge miss and several small misses can look equally bad even though we might just care about the big one.

### 6. Implement k-fold cross validation for k = 10.

Answer:

```{r}
# 10-fold cross validation for model_1 
set.seed(244)  
model_1_cv <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil,                 
                resamples = vfold_cv(train_data, v = 10, strata = "oil"),                  
                metrics = metric_set(mae, rsq))
```

```{r}
# 10-fold cross validation for model_2  
set.seed(253)  
model_2_cv <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil+school+invest,                 
                resamples = vfold_cv(train_data, v = 10, strata="oil"),                
                metrics = metric_set(mae, rsq))

```

### 7. Display evaluation metrics for your different models in a clean, organized way. This display should include both the estimated CV metric as well as its standard deviation.

Answer:

```{r}
model_1_cv %>%      
  collect_metrics()
```

```{r}
model_2_cv %>%      
  collect_metrics()
```

```{r}
# In sample MAE for model_1  
model_1 %>%      
  augment(new_data = train_data) %>%      
  mae(truth = gdpgrowth, estimate = .pred)
```

```{r}
# In sample MAE for model_2  
model_2 %>%     
  augment(new_data = train_data) %>%     
  mae(truth = gdpgrowth, estimate = .pred ) 
```

```{r}
model_1_cv %>%    
  unnest(.metrics) %>%      
  filter(.metric == "mae") %>%      
  summarize(sd(.estimate)) 
```

```{r}
model_2_cv %>%    
  unnest(.metrics) %>%      
  filter(.metric == "mae") %>%      
  summarize(sd(.estimate))
```

| Model   | In-Sample MAE | 10-fold CV MAE | Standard Deviations |
|---------|---------------|----------------|---------------------|
| model_1 | 1.39          | 1.44           | 0.42                |
| model_2 | 1.34          | 1.42           | 0.26                |

### 8. Try different values of k (the tuning parameter). At minimum, try k = n - 1 (LOOCV), and k = 5. Which value of k has the smallest CV error?

Answer: 5-fold CV gave the lowest mean MAE, so we will use k=5.

```{r}
# LOOCV  
model_1_loocv <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil,                 
                resamples = vfold_cv(train_data, v = nrow(train_data), strata="oil"),                 
                metrics = metric_set(mae))
```

```{r}
model_1_loocv %>%       
  collect_metrics()
```

```{r}
# 5-fold cv   
set.seed(244)  
model_1_cv5 <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil,                      
                resamples = vfold_cv(train_data, v = 5, strata="oil"),                     
                metrics = metric_set(mae, rsq))
```

```{r}
model_1_cv5 %>%      
  collect_metrics()
```

```{r}
model_2_loocv <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil+school+invest,                      
                resamples = vfold_cv(train_data, v = nrow(train_data), strata = "oil"),                 
                metrics = metric_set(mae))
```

```{r}
model_2_loocv %>%    
  collect_metrics()
```

```{r}
# 5-fold cv   
set.seed(253)  
model_2_cv5 <- lm_spec %>%      
  fit_resamples(gdpgrowth~oil+school+invest,                      
                resamples = vfold_cv(train_data, v = 5, strata = "oil"),                 
                metrics = metric_set(mae, rsq))
```

```{r}
model_2_cv5 %>%      
  collect_metrics()
```

### 9. Select your final model based on which one has the smallest CV error.

Answer: Model 2 has the smallest CV error compared to model 1. So, model_2 is my final model.

```{r}
# Refit the chosen model on all of train_data
final_model <- lm(gdpgrowth ~ oil + school + invest, data = train_data)

# Predict on the hold-out test set
test_preds <- predict(final_model, newdata = test_data)

# Compute the true out-of-sample MAE
test_mae <- mae_vec(
  truth    = test_data$gdpgrowth,
  estimate = test_preds
)

# Print it
test_mae

```

## LASSO

### 1. Write a brief introduction to LASSO which includes relevant mathematical notation.

Answer: LASSO is a penalized regression method that shrinks coefficient estimates and performs variable selection.

$\mathrm{RSS} + \lambda \sum_{j=1}^p |\beta_j|$

where lambda \>= 0 is a tuning parameter controlling the strength of the L1 penalty. As lambda grows more coefficients are shrinking towards 0 or exactly to 0 which yields a model that is easier to interpret and often stable in the presence of multicollinearity. LASSO's key advantages are it's built in variable selection and improved predictive performance via regularization and it's main drawback is potential bias in the non-zero coefficients due to shrinkage.

### 2. State which error metric you are using (i.e., CV MAE or CV MSE and state the chosen k-value).

Answer: I will me using CV MAE and k=10.

### 3. Identify at least p = 3 predictors X1, X2, ..., Xp for modeling the expected response E(Y) of one of your variables Y.

Answer: For the LASSO model, I will predict

$Y = gdpgrowth$

Using at least 3 predictors.

$X1 = gdp60$

$X2 = school$

$X3 = invest$

### 4. Compute 50 LASSO models for 50 different values of 𝜆 (the penalty tuning parameter), each with all p predictors. You can start out by trying 𝜆 in the range of -5 to 1 on the log10 scale. Plot the cross-validated (CV) error for each value of 𝜆.

Answer:

```{r}
# STEP 1: LASSO Algorithm & Model Specification 
# This is copied exactly from the R Code Notes for LASSO Section 
lasso_spec <- linear_reg() %>%   
  set_mode("regression") %>%   
  set_engine("glmnet") %>%   
  set_args(mixture = 1, penalty = tune()) 
```

```{r}
# STEP 2: Variable Recipe 
variable_recipe <- recipe(gdpgrowth ~ ., data = train_data) %>%   
  step_dummy(all_nominal_predictors())
```

```{r}
# STEP 3: Workflow Specification (Model + Recipe) 
# This is copied exactly from the R Code Notes for LASSO Section 
lasso_workflow <- workflow() %>%   
  add_recipe(variable_recipe) %>%   
  add_model(lasso_spec)
```

```{r}
# STEP 4: Estimate 50 LASSO models using # lambda values on a "grid" or range from 10^(-5) to 10^(-0.1). # Calculate the 10-fold CV MAE for each of the 50 models. 
set.seed(123) 
lasso_models<-lasso_workflow%>%   
  tune_grid(grid= grid_regular(penalty(range=c(-5,1)),levels= 50 ),
            resamples= vfold_cv(train_data, v= 10, strata = "oil" ),     
            metrics= metric_set(mae))
```

```{r}
autoplot(lasso_models)+ scale_x_log10() +  xlab(expression(lambda))
```

### 5. Based on your plot in the previous step, what value of lambda is best?

Answer: Based on the plot above, the lambda is best when it is equal to 0.25

```{r}
# Checking the approximation with the code 
best_penalty <- lasso_models %>%   
  select_best(metric = "mae") 
best_penalty
```

### 6. Pick the lambda associated with the parsimonious model.

Answer: 0.34

```{r}
# This is copied exactly from the R Code Notes for LASSO Section 
autoplot(lasso_models) +  
  scale_x_log10() +  
  xlab(expression(lambda)) +  
  geom_errorbar(data = collect_metrics(lasso_models), 
                aes(x = penalty, ymin = mean- std_err, 
                    ymax = mean +std_err), alpha = 0.5)
```

```{r}
parsimonious_penalty <- lasso_models %>%   
  select_by_one_std_err(metric = "mae", desc(penalty)) 
parsimonious_penalty
```

### 7. For your parsimonious model, how many and which predictors were kept?

Answer: For my parsimonious model, 4 variables were kept but rest of the other variables were shrunk exactly to zero. The variables that were kept are: gdp60, popgrowth, invest and inter_yes

```{r}
final_lasso <- lasso_workflow %>%   
  finalize_workflow(parameters = parsimonious_penalty) %>%   
  fit(data = train_data) 
final_lasso %>%   
  tidy()
```
